This is an example of article title

### 2023-12-28 paper.tex
Fixing up todos
Resolved some revisions with citations

%TODO: Review the abstract after reviewing the paper fully
Observational health continues to be a growing field in health informatics research as electronic health records (EHR), patient medical claims, and other ancilliary patient data source become more readily computable and accessible to researchers.
JuliaHealth is poised as an ecosystem to innovate within this area of research by bringing highly performant analytics approaches, composable solutions, and interoperable software that leverages prior state of the art. 
This paper will discuss the state of the art observational health research tools within the JuliaHealth ecosystem and how JuliaHealth is prepared to further research goals within this domain.

@@

% EXPLAIN: What is observational health research? Done
% EXPLAIN: What sort of innovations?
Over the past $10$ years, there have been several innovations within the observational health research community.  Observational health research is an area of study that leverages authentic health data to gain insights into the health system without interference. Data are collected by researchers as they follow groups of people over a specified period. Depending on the study, these groups may comprise healthy individuals, individuals with a specific disease, or those at high risk of developing a particular condition—such as individuals with a family history—while also taking into consideration the imperative of safeguarding their privacy. Observational studies can contribute to answering questions as what genes cause a specific disease to grow, patterns and trends for a particular disease, experiences of past patients, behaviours increasing the risk of developing a certain disease.

% EXPLAIN: Why is this a crucial innovation?
% EXPLAIN: What is meant by a transferable analysis
% TODO: Cite: J. M. Overhage, P. B. Ryan, C. G. Reich, A. G. Hartzema, and P. E. Stang, “Validation of a common data model for active safety surveillance research,” Journal of the American Medical Informatics Association, vol. 19, no. 1, pp. 54–60, 2012.
Arguably, the most crucial innovation that has emerged is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM) which was designed specifically to standardize observational health data to enable transferable analyses.
% TODO: Add citation from OHDSI website's collaborator statistics
% EXPLAIN: Why is mapping relevant here?
The importance of this model has become apparent within open science organizations like the OHDSI (Observational Health Data Science and Informatics) network where members across $100+$ countries with access to various health systems have been able to successfully map their patient data to the OMOP CDM.
% TODO: Add citation about successful collaborations
% EXPLAIN: What is a network study?
% TODO: Cite the quote "write once, run everywhere" 
% http://www.sun.com/smi/Press/sunflash/1996-01/sunflash.960123.10561.xml} 
This has led to successful collaboration relationships across countries where one research group can develop a network study and then disseminate the study across collaborators leading to the idea that you can "write once, run everywhere."

@@

% TODO: This paragraph should be centered on the core technology of data standards -- in particular, the OMOP CDM
% EXPLAIN: What is real world data? Use https://jacobzelko.com/10282021140730-real-world-evidence/
OMOP was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
What could arguably be considered the longest standing achievement of OMOP was the development of the OMOP Common Data Model (OMOP CDM) to house and ingest real world data.
% TODO: Combine the next two sentences together
% EXPLAIN: What does "person-centered" mean?
% EXPLAIN: What is a patient event? Could you give an example?
in a consistent format that has been optimized to assist in extracting meaningful population health information.
In brief, the OMOP Common Data Model (CDM) is a "person-centered" model that has standardized common events in a given patient encounter with database tables that represent the unique structures found in patient events.
% TODO: Develop this fragment into a full sentence
prioritizing data protection through the limiting of information that could endanger patient anonymity.
% EXPLAIN: What is a health system? Could you give an example?
This creates a longitudinal view of all healthcare-related events per person recorded in a singular health system.
% TODO: Make this more coherent to the rest of the paragraph
% TODO: Explicitly mention that although the model prescribes a database schema, it does not need a specific database type
The OMOP CDM itself does not require a specific technology to work with the data stored in this standard.

% TODO: This paragraph should be centered on the core technology of the concept of network studies
% TODO: Use this to address the notion of what "network studies" are
% TODO: Make this initially less OHDSI specific; pull from https://jacobzelko.com/05122023034151-network-macarthur
In particular, one collaboration mechanism that OHDSI has been able to operationalize meaningfully while working with observational health data is the notion of "network studies".
These studies, are designed to be transparent and encompasses research endeavors being explored by members of OHDSI comprehensive of relevant documentation, analysis code, results, and study protocols describing the study's scope and objectives.
OHDSI had its inception in the development of the Observational Medical Outcomes Partnership (OMOP) that was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
OHDSI was developed to focus on leveraging observational health data stored within the OMOP CDM.
Since current observational health data sources are aggregated at different levels and cannot be generalized to the overall population, the data is often incomplete, inaccurate, and prone to issues of bias, error handling, and multidimensional complexity. 
The goal of OHDSI is to produce reproducible, robust, and reliable evidence by developing standardized data formats and research protocol designs to enable large-scale collaborative research and improve the transparency and scalability of observational health research. 
% TODO: Carry HADES discussion into the paragraph talking about research tooling
One core innovation brought about by OHDSI in the space of network science is how, given the HADES tools and procedures established by OHDSI, 
% TODO: Make this more general such that it flows into the next paragraph on research tooling better
a research group can develop their analysis on their own OMOP CDM instance and then bundle together their study.
% EXPLAIN: What is a study package?
% EXPLAIN: What data assets?
Once this packaging is done, they can pass along the study package to another site to collaborate on the same analysis but on different data assets.
% EXPLAIN: What is a collaborator network?
Data confidentiality is preserved in this approach as data from individual participating sites remains exclusive to each site, with only aggregate results shared across a given collaborator network.
As such, no risk for re-identification comes about through this network collaboration and has ushered in the possibility for similar research approaches to be taken across the globe. \cite{ohdsi2019book}

% TODO: Develop this paragraph more about research software
As the OMOP CDM has become ubiquitous within observational health research, developing research software to handle and manage OMOP CDM data is necessary to meaningfully contribute within the observational health research space.
OHDSI supports the development of open-source software tools in a centralized ecosystem called HADES (Health Analytics Data-to-Evidence Suite) for particularly managing and working with OMOP CDM formatted data.
The HADES ecosystem has been in existence for $\approx 10$ years and has given rise to improvements in the quality and reliability of observational health research as well as motivating novel collaboration mechanisms.

@@

% TODO: Cite Discourse post on start of JuliaHealth
JuliaHealth is an open-science collective dedicated to bringing the Julia programming language to various avenues of health research domains including, but not limited to, observational health, medical imaging, standards interoperability and more.
It leverages the strengths of the Julia language to introduce new methods into these spaces, augment existing workflows, and provide a community structure for health researchers within the global Julia ecosystem.
% EXPLAIN: What is a subecosystem?
% EXPLAIN: How does having more subecosystems enable novel research to be done?
Eventually, JuliaHealth will host a variety of subecosystems and associated resources that will enable novel research to be conducted within Julia.

One area of the subecosystems within JuliaHealth that has matured significantly over the years has been the JuliaHealth observational health subecosystem (OHS).
It can be said that OHS had its inception in work done by Evans and Simonov with efforts such as \textit{FunOHDSI.jl}, \textit{OHDSICohortExpressions.jl}, and \textit{FunSQL.jl} around $2021$ \cite{kirill_simonov_2023_7705325}
% TODO: Cite FunOHDSI
% TODO: Cite OHDSICohortExpressions
% TODO: Cite OHDSI Symposium poster on JuliaHealth
% TODO: Register OHDSI Symposium poster to Zenodo
% TODO: Cite JuliaCon 2023 presentation
% TODO: Register JuliaCon 2023 presentation to Zenodo
Since then, Zelko et. al. have spearheaded most recent developments within OHS via the creation of a variety of tools and demonstrated their effectiveness in using these tools for novel observational health research endeavors \cite{zelko2022pilot}
% EXPLAIN: What does this mean?
Through such efforts, OHS is becoming further capable of executing novel observational health research.

@@

% TODO: Add a paragraph here about what is a phenotype definition
% TODO: Cite my paper here

% TODO: Segue this better from phenotype definitions
One of the premier tools within the OHDSI HADES ecosystem is the ATLAS Shiny application which was developed to develop patient phenotype definitions within observational health research contexts (amongst other use cases).
These phenotype definitions have varying levels of complexity where they can be very simple (such as finding all patients with one health condition or diagnosis) or grow to be extremely complex (defining controlled, uncontrolled, and indeterminate patient cohorts based on medications, lab results, etc.)
% TODO: Cite schema of the definition from GitHub
% EXPLAIN: Better introduce what SQL is
The computable phenotype definitions that ATLAS generates based on phenotype definitions are in a variety of serialization formats such as JSON or prepared SQL expressions available in multiple SQL flavors.

% TODO: Cite package
% TODO: Reframe this to more say how ATLAS can be used to build cohorts
% TODO: Explain how this package to accomplish similar tasks
The Julia package, OHDSICohortExpressions.jl, exists specifically to ingest these serialized expressions of computable phenotype definitions and create patient cohorts that match the given criterion in the definition.
% TODO: Change this list into an enumeration
How this tool works is to 1) read and verify the serialized computable phenotype definition 2) reinterpret the serialization into the FunSQL.jl Domain Specific Language 3) return a prepared SQL statement against a given SQL database that queries an OMOP CDM database for the desired patient population(s) of interest.
% EXPLAIN: What is a prepared statement?
% EXPLAIN: What is a cohort in this context?
Once these prepared statements are made, researchers can execute the generated SQL against their OMOP CDM instance to generate cohorts for future analyses.

@@

% TODO: Reword this sentence
Although OHDSICohortExpressions.jl is sufficient to translate existing phenotype definitions generated in ATLAS to executable queries, oftentimes, there is a tricky problem of how to best iterate on phenotype definitions. \cite{zelkoDevelopingRobustComputable2023}
% EXPLAIN: Better contextualize why this is important -- why not just create a new phenotype definition?
Often, one has to take an existing definition and either revise to a more narrow cohort or write manual code to build more niche cohorts after an initial definition is created.
% EXPLAIN: How so?
% TODO: Review this sentence; maybe cut it
This refactoring of a phenotype definition can lead to conflating ideas within conversation between expert clinical reviewers, analysts, and health researchers.
% TODO: Move this to the previous section on phenotype definitions
This broadly can be the notions of what we are exploring (the phenotype definition; often the responsibility of a clinical reviewer), what is possible to compute upon (the computable phenotype definition; often the responsibility of an analyst), and the goals of the overall study.

% TODO: Present this more as a general introduction to what tools are here for this
% TODO: Talk about composition with DataFrames or Tidier
OMOPCDMCohortCreator.jl allows one to iteratively build a patient population quickly and easily.
One can use patient pools developed from pre-existing cohorts defined off of phenotype definitions as a starting point in exploring patients or create new patient populations quickly without the need for a well-defined phenotype definition.
% TODO: Mention the ability to satisfy research requests
This can enable analysts to more quickly understand what is inside their data assets, do "quick" analysis of a particular smaller research question, and enable a more distinct separation of work between a clinical expert reviewer and analyst teams.

% TODO: Mention capr

@@

% TODO: Mention JuliaCon Birds of a Feather discussion
One of the biggest styming factors within not only the Julia ecosystem, but observational health research in general has been the difficulty of working with different databases.
Although the OMOP CDM has been great for standardizing \textit{how} real world data is handled, there is a lack of consensus on \textit{what} is needed to handle these data assets.
% TODO: Make this less verbose
For a variety of reasons, different groups will choose different database architectures (perhaps PostgreSQL for improved permission support, DuckDB for performance, or even SQLite for ease of use) but all are subtly different enough to create headaches in seamlessly working across different research sites.

% TODO: Change the phrasing of this
The way that the Julia ecosystem handles this is via the creation of a common database interface called the DBInterface.
% EXPLAIN: What is dispatch
Specific Julia database packages can implement dispatches for this interace allowing users to have a much better experience working across databases.
In spite of such benefits, for some users (and especially within the context of observational health studies), it can beneficial to have a more prescriptive interface where one just needs to pass perhaps a password, username, database name, and schema.

% TODO: Better introduce this
For that reason, DBConnector.jl was created.
% EXPLAIN: Contextualize this a bit better
Although this is not an explicit JuliaHealth package, it was a pain point that was experienced throughout the process of observational health research across multiple partner sites.
The promise of DBConnector.jl is that with one line command, users can connect simply, and in a common (albeit opinionated) manner, to databases.
DBConnector unifies Julia database packages that implement the DBInterface to achieve this level of simplified connection making.
DBConnector.jl was inspired by the package \textit{DatabaseConnector} (Link: https://github.com/OHDSI/DatabaseConnector)

% TODO: Combine with the previous paragraph and state that there are two approaches
% TODO: Make explicit DBConnector's path
% TODO: Make explicit the path of choosing general packages
Finally, if someone needs more fine-grained control over a particular database connection, they can choose to eschew this package.
Instead, they can switch to the specific Julia database package that would give them the additional control.
Although they'll lose the simplicity of this package, core required functionality is not lost when working with databases.

% TODO: Review if this section is actually needed
\subsection{Ecosystem Interoperability}

@@

% TODO: Reference code repository
Using the JuliaHealth tools developed, we enumerate an example observational health study workflow that would be used in practice.\footnotemark
For this paper, we used two datasets: a synthetic patient dataset called \textit{Eunomia} and the MIMIC III patient dataset which had been converted into the OMOP CDM to illustrate the workflow provided by the Observational Health sub-ecosystem.
Reproducing the workflow using the \textit{Eunomia} dataset is made readily available at the repository here: \cite{schuemieEunomia2023}

% TODO: Add link to code repository
\footnotetext{The code snippets presented in the body of the paper may omit certain intermediate steps. 
Please review the associated code repository for a full working tutorial.}

@@

@@

\subsection{Bridging between OHDSI and JuliaHealth Communities}

@@

After having a strep throat phenotype definition defined using ATLAS, we can then execute the resulting computable phenotype definition against the \textit{Eunomia} database.

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
using JSON3 
using OHDSICohortExpressions

@@

for query in split(sql, ";")[1:end-1]
  DBI.execute(conn, query)
end
\end{minted}

A quick query of the cohort table (where the resulting subjects were stored) shows $\approx 1600$ from the dataset match this criteria.

@@

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
C = GetCohortSubjects(1, conn)
rename!(C, :subject_id => :person_id)

@@

C_age_group = occ.GetPatientAgeGroup(
  C.person_id, 
  conn; 
)
\end{minted}

@@

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
import DataFrames as DF

@@

C_groups = DF.combine(C_groups, DF.nrow => :count)
\end{minted}

Using this analytical approach, tabulation can be quickly conducted across patients.
Additionally, throughout the process, one can perform relevant analysis methods.
Here, we'll conduct a crude prevalence calculation across the patient cohort.

\begin{table}[!ht]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|}
    \hline
        Race & Gender & Age Group & Cohort & Total & Prev. (\%) \\ \hline
        B/AA & F & 60 - 64 & 13 & 21 & 61.90 \\ \hline
        W & F & 50 - 54 & 59 & 111 & 53.15 \\ \hline
        W & F & 55 - 59 & 62 & 97 & 63.91 \\ \hline
        A & M & 80 - 84 & 6 & 10 & 60.00 \\ \hline
        UNK & F & 60 - 64 & 29 & 39 & 74.35 \\ \hline
        B/AA & M & 65 - 69 & 12 & 18 & 66.66 \\ \hline
    \end{tabular}
\end{table}

@@

As seen in Table 1 and Table 2, informative results can be rapidly generated by this workflow relevant to observational health research.
Although the values shown in Table 2 are synthetic and the crude prevalence values themselves are meaningless, we can take a this approach and apply it directly to the MIMIC III dataset.
In Table 3, we can calculate across this dataset meaningful statistics that could later be further analyzed and drive further questioning (such as ...).

@@

    1. Modularization of code and chaining
    2. Distributed computing

\subsubsection{Modularization of Code and Chaining}

An exciting emergent aspect of the JuliaHealth ecosystem that came about as a result of developing OMOPCDMCohortCreator.jl was the prospect of radical modularization and chaining of package functionalities together.
For example, taking our workflow example from the earlier section where we characterized patients by race, gender, and age group, we can do a few "tricks" to modularize these characterization functions as follows:

>>>>

% TODO::HARD Review the abstract after reviewing the paper fully
Observational health continues to be a growing field in health informatics research as electronic health records (EHR), patient medical claims, and other ancilliary patient data source become more readily computable and accessible to researchers.
JuliaHealth is poised as an ecosystem to innovate within this area of research by bringing highly performant analytics approaches, composable solutions, and interoperable software that leverages prior state of the art. 
This paper will discuss the state of the art observational health research tools within the JuliaHealth ecosystem and how JuliaHealth is prepared to further research goals within this domain.

@@

% EXPLAIN::EASY What is observational health research?
% EXPLAIN::MEDIUM What sort of innovations?
Over the past $10$ years, there have been several innovations within the observational health research community.  Observational health research is an area of study that leverages authentic health data to gain insights into the health system without interference. Data are collected by researchers as they follow groups of people over a specified period. Depending on the study, these groups may comprise healthy individuals, individuals with a specific disease, or those at high risk of developing a particular condition—such as individuals with a family history—while also taking into consideration the imperative of safeguarding their privacy. Observational studies can contribute to answering questions as what genes cause a specific disease to grow, patterns and trends for a particular disease, experiences of past patients, behaviours increasing the risk of developing a certain disease.

% EXPLAIN::MEDIUM Why is this a crucial innovation?
% EXPLAIN::MEDIUM What is meant by a transferable analysis?
Arguably, the most crucial innovation that has emerged is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM) which was designed specifically to standardize observational health data to enable transferable analyses. \cite{overhage2012validation}
% EXPLAIN::MEDIUM Why is mapping relevant here?
The importance of this model has become apparent within open science organizations like the OHDSI (Observational Health Data Science and Informatics) network where members across $100+$ countries \cite{sachsonOurJourney2023} with access to various health systems have been able to successfully map their patient data to the OMOP CDM.
% TODO::EASY Add citation about successful collaborations
% EXPLAIN::EASY What is a network study?
% TODO::EASY Cite the quote "write once, run everywhere" 
% http://www.sun.com/smi/Press/sunflash/1996-01/sunflash.960123.10561.xml} 
This has led to successful collaboration relationships across countries where one research group can develop a network study and then disseminate the study across collaborators leading to the idea that you can "write once, run everywhere."

@@

% TODO::MEDIUM This paragraph should be centered on the core technology of data standards -- in particular, the OMOP CDM
% EXPLAIN::MEDIUM What is real world data? Use https://jacobzelko.com/10282021140730-real-world-evidence/
OMOP was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
What could arguably be considered the longest standing achievement of OMOP was the development of the OMOP Common Data Model (OMOP CDM) to house and ingest real world data.
% TODO::EASY Combine the next two sentences together
% EXPLAIN::EASY What does "person-centered" mean?
% EXPLAIN::MEDIUM What is a patient event? Could you give an example?
in a consistent format that has been optimized to assist in extracting meaningful population health information.
In brief, the OMOP Common Data Model (CDM) is a "person-centered" model that has standardized common events in a given patient encounter with database tables that represent the unique structures found in patient events.
% TODO::EASY Develop this fragment into a full sentence
prioritizing data protection through the limiting of information that could endanger patient anonymity.
% EXPLAIN:: What is a health system? Could you give an example?
This creates a longitudinal view of all healthcare-related events per person recorded in a singular health system.
% TODO::EASY Make this more coherent to the rest of the paragraph
% TODO::EASY Explicitly mention that although the model prescribes a database schema, it does not need a specific database type
The OMOP CDM itself does not require a specific technology to work with the data stored in this standard.

% TODO::MEDIUM This paragraph should be centered on the core technology of the concept of network studies
% TODO::MEDIUM Use this to address the notion of what "network studies" are
% TODO::MEDIUM Make this initially less OHDSI specific; pull from https://jacobzelko.com/05122023034151-network-macarthur
In particular, one collaboration mechanism that OHDSI has been able to operationalize meaningfully while working with observational health data is the notion of "network studies".
These studies, are designed to be transparent and encompasses research endeavors being explored by members of OHDSI comprehensive of relevant documentation, analysis code, results, and study protocols describing the study's scope and objectives.
OHDSI had its inception in the development of the Observational Medical Outcomes Partnership (OMOP) that was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
OHDSI was developed to focus on leveraging observational health data stored within the OMOP CDM.
Since current observational health data sources are aggregated at different levels and cannot be generalized to the overall population, the data is often incomplete, inaccurate, and prone to issues of bias, error handling, and multidimensional complexity. 
The goal of OHDSI is to produce reproducible, robust, and reliable evidence by developing standardized data formats and research protocol designs to enable large-scale collaborative research and improve the transparency and scalability of observational health research. 
% TODO::MEDIUM Carry HADES discussion into the paragraph talking about research tooling
One core innovation brought about by OHDSI in the space of network science is how, given the HADES tools and procedures established by OHDSI, 
% TODO::EASY Make this more general such that it flows into the next paragraph on research tooling better
a research group can develop their analysis on their own OMOP CDM instance and then bundle together their study.
% EXPLAIN::EASY What is a study package?
% EXPLAIN::EASY What data assets?
Once this packaging is done, they can pass along the study package to another site to collaborate on the same analysis but on different data assets.
% EXPLAIN::MEDIUM What is a collaborator network?
Data confidentiality is preserved in this approach as data from individual participating sites remains exclusive to each site, with only aggregate results shared across a given collaborator network.
As such, no risk for re-identification comes about through this network collaboration and has ushered in the possibility for similar research approaches to be taken across the globe. \cite{ohdsi2019book}

% TODO::HARD Develop this paragraph more about research software
As the OMOP CDM has become ubiquitous within observational health research, developing research software to handle and manage OMOP CDM data is necessary to meaningfully contribute within the observational health research space.
OHDSI supports the development of open-source software tools in a centralized ecosystem called HADES (Health Analytics Data-to-Evidence Suite) for particularly managing and working with OMOP CDM formatted data.
The HADES ecosystem has been in existence for $\approx 10$ years and has given rise to improvements in the quality and reliability of observational health research as well as motivating novel collaboration mechanisms.

@@

JuliaHealth is an open-science collective dedicated to bringing the Julia programming language to various avenues of health research domains including, but not limited to, observational health, medical imaging, standards interoperability and more.\cite{aluthgeAnnouncingJuliaHealthOrganization2020}
It leverages the strengths of the Julia language to introduce new methods into these spaces, augment existing workflows, and provide a community structure for health researchers within the global Julia ecosystem.
% EXPLAIN::EASY What is a subecosystem?
% EXPLAIN:: How does having more subecosystems enable novel research to be done?
Eventually, JuliaHealth will host a variety of subecosystems and associated resources that will enable novel research to be conducted within Julia.

One area of the subecosystems within JuliaHealth that has matured significantly over the years has been the JuliaHealth observational health subecosystem (OHS).
It can be said that OHS had its inception in work done by Evans and Simonov with efforts such as \textit{FunOHDSI.jl} \cite{evansFunOHDSIJl2022}, \textit{OHDSICohortExpressions.jl} \cite{evansOHDSICohortExpressionsJl2023}, and \textit{FunSQL.jl} around $2021$ \cite{kirill_simonov_2023_7705325}
% TODO::EASY Cite OHDSI Symposium poster on JuliaHealth
% TODO::HARD Register OHDSI Symposium poster to Zenodo
% TODO::EASY Cite JuliaCon 2023 presentation
% TODO::HARD Register JuliaCon 2023 presentation to Zenodo
Since then, Zelko et. al. have spearheaded most recent developments within OHS via the creation of a variety of tools and demonstrated their effectiveness in using these tools for novel observational health research endeavors \cite{zelko2022pilot}
% EXPLAIN::MEDIUM What does this mean?
Through such efforts, OHS is becoming further capable of executing novel observational health research.

@@

% TODO::MEDIUM Add a paragraph here about what is a phenotype definition
% TODO::EASY Cite my paper here

% TODO::EASY Segue this better from phenotype definitions
One of the premier tools within the OHDSI HADES ecosystem is the ATLAS Shiny application which was developed to develop patient phenotype definitions within observational health research contexts (amongst other use cases).
These phenotype definitions have varying levels of complexity where they can be very simple (such as finding all patients with one health condition or diagnosis) or grow to be extremely complex (defining controlled, uncontrolled, and indeterminate patient cohorts based on medications, lab results, etc.)
% TODO::EASY Create footnote about schema
% TODO::EASY Add reference to schema in footnote
% TODO::EASY Cite reference schema from GitHub in footnote; \cite{madrilATLASJSONSchema2022}
% EXPLAIN::EASY Better introduce what SQL is
The computable phenotype definitions that ATLAS generates based on phenotype definitions are in a variety of serialization formats such as JSON or prepared SQL expressions available in multiple SQL flavors.

% TODO::MEDIUM Reframe this to more say how ATLAS can be used to build cohorts
% TODO::EASY Explain how this package accomplishes similar tasks
The Julia package, OHDSICohortExpressions.jl, exists specifically to ingest these serialized expressions of computable phenotype definitions and create patient cohorts that match the given criterion in the definition. \cite{evansOHDSICohortExpressionsJl2023}
% TODO::MEDIUM Turn all items into complete thoughts in list
How this tool works is to:

\item read and verify the serialized computable phenotype definition

\item reinterpret the serialization into the FunSQL.jl Domain Specific Language

\item return a prepared SQL statement against a given SQL database that queries an OMOP CDM database for the desired patient population(s) of interest.

% EXPLAIN::MEDIUM What is a prepared statement?
% EXPLAIN::MEDIUM What is a cohort in this context?
Once these prepared statements are made, researchers can execute the generated SQL against their OMOP CDM instance to generate cohorts for future analyses.

@@

% TODO::EASY Reword this sentence
Although OHDSICohortExpressions.jl is sufficient to translate existing phenotype definitions generated in ATLAS to executable queries, oftentimes, there is a tricky problem of how to best iterate on phenotype definitions. \cite{zelkoDevelopingRobustComputable2023}
% EXPLAIN:: Better contextualize why this is important -- why not just create a new phenotype definition?
Often, one has to take an existing definition and either revise to a more narrow cohort or write manual code to build more niche cohorts after an initial definition is created.
% EXPLAIN::MEDIUM How so?
% TODO::MEDIUM Review this sentence; maybe cut it
This refactoring of a phenotype definition can lead to conflating ideas within conversation between expert clinical reviewers, analysts, and health researchers.
% TODO::EASY Move this to the previous section on phenotype definitions
This broadly can be the notions of what we are exploring (the phenotype definition; often the responsibility of a clinical reviewer), what is possible to compute upon (the computable phenotype definition; often the responsibility of an analyst), and the goals of the overall study.

% TODO::EASY Present this more as a general introduction to what tools are here for this
% TODO::EASY Talk about composition with DataFrames or Tidier
OMOPCDMCohortCreator.jl allows one to iteratively build a patient population quickly and easily.
One can use patient pools developed from pre-existing cohorts defined off of phenotype definitions as a starting point in exploring patients or create new patient populations quickly without the need for a well-defined phenotype definition.
% TODO::MEDIUM Mention the ability to satisfy research requests
This can enable analysts to more quickly understand what is inside their data assets, do "quick" analysis of a particular smaller research question, and enable a more distinct separation of work between a clinical expert reviewer and analyst teams.

@@

One of the biggest styming factors within not only the Julia ecosystem, but observational health research in general has been the difficulty of working with different databases.\cite{zelko2023julia}
Although the OMOP CDM has been great for standardizing \textit{how} real world data is handled, there is a lack of consensus on \textit{what} is needed to handle these data assets.
% TODO::EASY Make this less verbose
For a variety of reasons, different groups will choose different database architectures (perhaps PostgreSQL for improved permission support, DuckDB for performance, or even SQLite for ease of use) but all are subtly different enough to create headaches in seamlessly working across different research sites.

% TODO::EASY Change the phrasing of this
The way that the Julia ecosystem handles this is via the creation of a common database interface called the DBInterface.
% EXPLAIN:: What is dispatch
Specific Julia database packages can implement dispatches for this interace allowing users to have a much better experience working across databases.
In spite of such benefits, for some users (and especially within the context of observational health studies), it can beneficial to have a more prescriptive interface where one just needs to pass perhaps a password, username, database name, and schema.

% TODO::MEDIUM Better introduce this
For that reason, DBConnector.jl was created.
% EXPLAIN::MEDIUM Contextualize this a bit better
Although this is not an explicit JuliaHealth package, it was a pain point that was experienced throughout the process of observational health research across multiple partner sites.
The promise of DBConnector.jl is that with one line command, users can connect simply, and in a common (albeit opinionated) manner, to databases.
DBConnector unifies Julia database packages that implement the DBInterface to achieve this level of simplified connection making.
DBConnector.jl was inspired by the package \textit{DatabaseConnector} (Link: https://github.com/OHDSI/DatabaseConnector)

% TODO::MEDIUM Combine with the previous paragraph and state that there are two approaches
% TODO::EASY Make explicit DBConnector's path
% TODO::EASY Make explicit the path of choosing general packages
Finally, if someone needs more fine-grained control over a particular database connection, they can choose to eschew this package.
Instead, they can switch to the specific Julia database package that would give them the additional control.
Although they'll lose the simplicity of this package, core required functionality is not lost when working with databases.

% TODO::HARD Review if this section is actually needed
\subsection{Ecosystem Interoperability}

@@

% TODO::EASY Reference code repository
Using the JuliaHealth tools developed, we enumerate an example observational health study workflow that would be used in practice.\footnotemark
% EXPLAIN::EASY In enumeration talk about what Eunomia is 
% EXPLAIN::EASY In enumeration discuss MIMIC III

For this paper, we used two datasets: a synthetic patient dataset called \textit{Eunomia} and the MIMIC III patient dataset which had been converted into the OMOP CDM to illustrate the workflow provided by the Observational Health sub-ecosystem.\cite{physionet, hripcsakCharacterizingTreatmentPathways2016, johnson2016mimic, johnson2018mimic}

\item \textbf{Eunomia} -

\item \textbf{MIMIC III} -

% TODO::EASY Mention in footnote explicitly that reproducing workflow in Eunomia is possible
% TODO::EASY Mention in footnote that explicitly reproducing workflow in MIMIC III is possible but not readily accessible
Reproducing the workflow using the \textit{Eunomia} dataset is made readily available at the repository here: \cite{schuemieEunomia2023}

% TODO::EASY Add link to code repository
\footnotetext{The code snippets presented in the body of the paper may omit certain intermediate steps. 
Please review the associated code repository for a full working tutorial.}

@@

% EXPLAIN::EASY What is this doing here?
% TODO::HARD Update DBConnector
% TODO::MEDIUM Maybe remove package imports and mention them within the caption instead

@@

% TODO::HARD Review if this subsection is needed
\subsection{Bridging between OHDSI and JuliaHealth Communities}

@@

% EXPLAIN:: What is this doing here?
After having a strep throat phenotype definition defined using ATLAS, we can then execute the resulting computable phenotype definition against the \textit{Eunomia} database.

% TODO::HARD Maybe remove package imports and mention them within the caption instead
\begin{listing}[!ht]
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
using JSON3 
using OHDSICohortExpressions

@@

for query in split(sql, ";")[1:end-1]
  DBI.execute(conn, query)
end
\end{minted}
% TODO::EASY Add a caption
\caption{}
% TODO::EASY Add a label
\label{listing:}
\end{listing}

% TODO::EASY Move this to a caption
A quick query of the cohort table (where the resulting subjects were stored) shows $\approx 1600$ from the dataset match this criteria.

@@

% EXPLAIN:: What is this section doing?

% TODO::EASY Maybe remove package imports and mention them within the caption instead
\begin{listing}[!ht]
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
C = GetCohortSubjects(1, conn)
rename!(C, :subject_id => :person_id)

@@

C_age_group = occ.GetPatientAgeGroup(
  C.person_id, 
  conn; 
)
\end{minted}
% TODO::EASY Add a caption
\caption{}
% TODO::EASY Add a label
\label{listing:}
\end{listing}

@@

% EXPLAIN::MEDIUM What is this section doing?

% TODO::EASY Maybe remove package imports and mention them within the caption instead

\begin{listing}[!ht]
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
import DataFrames as DF

@@

C_groups = DF.combine(C_groups, DF.nrow => :count)
\end{minted}
% TODO::EASY Add a caption
\caption{}
% TODO::EASY Add a label
\label{listing:}
\end{listing}

Using this analytical approach, tabulation can be quickly conducted across patients.

% EXPLAIN::EASY Move this to the caption of the table
Additionally, throughout the process, one can perform relevant analysis methods.
Here, we'll conduct a crude prevalence calculation across the patient cohort.

% EXPLAIN::MEDIUM What is this section doing?

\begin{table}[!ht]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|}
    \hline
        Race & Gender & Age Group & Cohort & Total & Prev. (\%) \\ \hline
        B/AA & F & 60 - 64 & 13 & 21 & 61.90 \\ \hline
        W & F & 50 - 54 & 59 & 111 & 53.15 \\ \hline
        W & F & 55 - 59 & 62 & 97 & 63.91 \\ \hline
        A & M & 80 - 84 & 6 & 10 & 60.00 \\ \hline
        UNK & F & 60 - 64 & 29 & 39 & 74.35 \\ \hline
        B/AA & M & 65 - 69 & 12 & 18 & 66.66 \\ \hline
    \end{tabular}
    % TODO::EASY Add a caption
    \caption{}
    % TODO::EASY Add a label
    \label{table:}
\end{table}

@@

% TODO::HARD Create table for MIMIC III results
% TODO::EASY Create reference to Eunomia table
As seen in Table 1, informative results can be rapidly generated by this workflow relevant to observational health research.
Although the values shown in Table 2 are synthetic and the crude prevalence values themselves are meaningless, we can take a this approach and apply it directly to the MIMIC III dataset.
In Table 3, we can calculate across this dataset meaningful statistics that could later be further analyzed and drive further questioning (such as ...).

@@

% EXPLAIN::MEDIUM What is the overview of this section?
% EXPLAIN::MEDIUM What does each topic do?
\begin{enumerate}

\item Modularization of code and chaining

\item Distributed computing

\item Automatic generation of SQL from health packages

% TODO::EASY Rename section
\subsubsection{Modularization of Code and Chaining}
